Rising from the Ashes: A Metallicity-Dependent Star Formation Gap Splits the Milky Way's $\alpha$-Sequences

### 2025-02-13 paper/ms.tex
Update discussion intro par and first section

\section{Discussion}\label{sec:discussion}
We have demonstrated that a bimodal structure in the elemental abundance plane is associated with a brief quiescent phase in our idealized merger simulations. We now construct a possible framework for understanding the formation of the bimodality, discuss its relation to prior work, and then discuss predictions and connections to observations.

\subsection{Quiescence Leads to Bimodality}\label{ssec:formqui}
We executed a series of idealized merger simulations in which we modified the starting radius and velocity by $\pm10\%$ and the circularity by $\pm0.1$, for a total of $27$ simulations. The central and satellite galaxies, which are meant to resemble the Milky Way and GSE at $z\sim2$, are otherwise identical across the simulations. Some of these simulations induce a bimodality, while others do not. We have examined a representative of each scenario in detail.

We described a series of results in Section~\ref{sec:results} which give a natural explanation for the separation between the high- and low-$\alpha$ sequences in our simulations. First, a merger induces a starburst in the simulation. If this starburst does not induce a quiescent period, then the median \MgFe{} of the gas briefly increases before continuing along the typical evolutionary path (right panel of Figure~\ref{fig:SFR_alpha}). However, if there is a quiescent period, the median \MgFe{} drops for the duration of the quiescent period before recovering to the typical path (left panel of Figure~\ref{fig:SFR_alpha}).

This drop can be understood in terms of the mechanics of the galaxy formation model as described in Section~\ref{ssec:gfm}. When a star particle forms, it only takes $\sim40\,\Myr$ before the last Type II SN has gone off and its material enrichment is dominated by Fe from Type Ia SNe. As a result, during the quiescent phase which lasts a few hundred Myr, there is relatively little $\alpha$-element production. On the other hand, star particles are still able to generate Fe through Type Ia SNe which have a longer period of influence. The fact that the quiescent period is preceded by a starburst is important. Even though star particles produce Fe through the Type Ia channel for Gyr, younger star particles still produce much more Fe than older star particles. In the TNG model, the number of Type Ia SNe decays with time as $(t/\tau_8)^{-s}$, where $\tau_8=40\,\Myr$ is the lifetime of an $8\,\Msun$ star and $s=1.12$. Therefore, the burst of star formation at $t\sim2.25\,\Gyr$ is able to generously produce Fe starting at $2.5\,\Gyr$, when is indeed when the drop in \MgFe{} is the most steep.

>>>>

\section{Discussion}\label{sec:discussion}
\edit1{We have investigated the formation of an $\alpha$-element bimodality in the Milky Way through a series of idealized merger simulations. Our key finding is that a metallicity-dependent quiescent period in star formation can lead to a bimodal distribution in \alphaFe{} at specific \FeH{} values. This scenario does not necessarily require a global quenching period. We now discuss the details of this mechanism, its connection to high-redshift observations and cosmological simulations, compare with some explanations in the literature, and explore its observational implications and directions for future work.}

\subsection{\FeH{}-dependent Quiescence Leads to Bimodality}\label{ssec:formqui}
\edit1{We executed a series of idealized merger simulations in which we modified the starting radius and velocity by $\pm10\%$ and the circularity by $\pm0.1$, for a total of $27$ simulations. The central and satellite galaxies, which are meant to resemble the Milky Way and GSE at $z\sim2$, are otherwise identical across the simulations. Some of these simulations induce a bimodality, while others do not. We have examined a representative of each scenario in detail.}

\edit1{The key driver of bimodality in our simulations is the presence of a metallicity-dependent quiescent period. This is shown most clearly in Figure~\ref{fig:before_after}, where the abundance plane is split into stars which form before and after a gap in the $\FeH=0$ SFR. Stars which form before and after the gap populate the high- and low-$\alpha$ sequences, respectively, with minimal overlap. No gap and no separation between the sequences is seen in the unimodal case (Figure~\ref{fig:before_after_uni}).}

\edit1{This perspective is bolstered by examining the distributions of \MgFe{} at $\FeH=0$ for the full simulation suite (Figure~\ref{fig:all_hist}). We order these simulations alphabetically using the bimodality score ($\mathcal{B}$, see Section~\ref{ssec:bim_score}). Simulations~r through aa are considered bimodal based on a visual inspection, and the location of their trough \MgFe{} (minimium of the distribution) is indicated with a vertical line.}

\edit1{The formation of these bimodal populations can be understood from Figure~\ref{fig:all_scatter}, which shows a scatter plot of \MgFe{} and formation time in the same order, with the trough \MgFe{} indicated with a horizontal line. Here, we can see that bimodal simulations tend to have a gap shortly after their respective mergers ($\sim2$--$3\,\Gyr$, depending on the orbital configuration), lying at the position of the trough \MgFe{}.}

\edit1{There are two exceptions in the bimodal cases. First, simulation~t does not have a complete gap, although the number of stars forming does still drop. This indicates that the complete absence of star formation is not necessary, but rather a reduction in the \FeH{}-dependent SFR may be sufficient if the \alphaFe{} ratio is declining fast enough. Second, simulation~w exhibits some irregular behavior, with star formation switching between high- and low-$\alpha$ multiple times.}

\edit1{There are a few unimodal simulations that have gaps -- e.g., d, f, g, h, and o. However, in these cases there is very little star formation at $\FeH=0$ before the gap, and so in these cases there is not a distinct high-$\alpha$ mode that can form.}

\edit1{Overall, these simulations indicate that a gap in star formation at a specific metallicity can lead to a bimodality in the conditional \alphaFe{} distribution at that metallicity. Determining precisely the mechanism behind generating these gaps is beyond the scope of this work, but we speculate briefly in Section~\ref{ssec:future_work}.}
